---
title: "Learning R - Syntax"
author: Florent Dufour
date: 2017 - 2019
output: 
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: simplex
---

# Variables and operators

## Assign a variable

These 3 assignements are correct:

```{r}
variableA = 2
variable.a = 2
a <-2
```

## Data types

The different types:
```{r}
print(class(4))                      # numeric
print(class(4L))                     # integer
print(class(TRUE))                   # logical
print(class(T))                      # logical
print(class(FALSE))                  # logical
print(class(F))                      # logical
print(class(1 +4i))                  # complex
print(class("Pancake"))              # character
print(class(charToRaw("Blueberry"))) # raw
```

The operation with types:
```{r}
# Check the type
num = 2
int = 6L
is.integer(num)
is.integer(int)
is.integer(num)
is.numeric(int)

# Convert types (if possible)
as.integer(num)
```

## Arithmetic operators

```{r}
sprintf("1 + 2 = %d", 1+2)
sprintf("2 - 3 = %d", 2-3)
sprintf("3 * 4 = %d", 3*4)
sprintf("4 / 5 = %1.3f", 4/5)
sprintf("5 %% 4 = %d", 5%%4)
sprintf("4^2 = %d", 4^2)
```

## Relational operators

```{r}
sprintf("4 == 5: %s", 4==5)
sprintf("4 != 5: %s", 4!=5)
sprintf("4 > 5: %s", 4>5)
sprintf("4 < 5: %s", 4<5)
sprintf("4 >= 5: %s", 4>=5)
sprintf("4 <= 5: %s", 4<=5)
```

## Vectors

They store multiple variables:
```{r}
numbers = c(3,2,0,1,8)
numbers[1] # WARNING: Vectors start at 1
length(numbers)
numbers[length(numbers)] # last
numbers[-1] # We don't want the first one
numbers[c(1,2)] # Two first values; again, start at 1
numbers[5] = 1 # Replace the 5th value by 1
numbers[c(1,2)] = c(4,5) # Replace two forst values
sort(numbers) # sort ascending
sort(numbers, decreasing=T) # sort decreasing
```

Generate sequences:
```{r}
# Range
oneToTen = 1:10

# Sequence
add3 = seq(from=3, to=27, by=3)
evens = seq(from=2, by=2, length.out=10)

# Repeat
rep(x=2, times=5, each=2)
rep(c(1, 2, 3), times=2, each=2)
```

Methods on vectors:
```{r}
sprintf("4 in evens %s: ", 4 %in% evens)

# Array of boolean; isEven?
oneTo20 = c(1:20)
isEven = oneTo20 %% 2 == 0
print(isEven)

# Create an array of evens
justEvens = oneTo20[oneTo20 %% 2 == 0]
print(justEvens)
```

## Strings

```{r}
str1 = "This is a string"

# Methods on strings
nchar(str1)

# Compare strings
sprintf("pancake > blueberry %s", "dog" > "blueberry")
sprintf("pancake == blueberry %s", "dog" == "blueberry")

# Combine strings
str2 = paste("Blueberry", "pancake", sep="+")

# Delete strings
str3 = "This is also a string"
substr(x=str3, start=1, stop=7)

# Substitute (pattern handles regex)
## First match
sub(pattern="pancake", replacement="jam", x=str2)
## All matches
gsub(pattern="Egg", replacement="Chicken", x="Egg and Egg")

# Split a string in vector
strVect = strsplit("A dog ran fast", " ")
```

## Factors

- Use when limited amount of values (strings or int).
- Notion of levels: A factor contains levels that are possible value

```{r}

# A factor
direction = c("up", "left", "right", "down", "left")
factorDirection = factor(direction)
is.factor(factorDirection)

# And its levels
levels(x=factorDirection)
```

Define the levels and their order:
```{r}
daysOfTheWeek = c(
  "Monday",
  "Tuesday",
  "Wednesday",
  "Thursday",
  "Friday",
  "Saturday",
  "Sunday"
)

workingdDays = c("Wednesday", "Monday", "Tuesday")

workingDaysFactors = factor(x=workingdDays, levels=daysOfTheWeek, ordered=TRUE)
workingDaysFactors # 3 days and the 7 levels
```

## Data frames

Is a table that contains any type of data in columns.
- A row = A record
- A column =  A variable

```{r}
customerData = data.frame(
  name = c("Tom", "Sally", "Sue"),
  age = c(43,28,35),
  stringsAsFactors = T
)

# Dimension
dim(customerData)
# Data in row 1, column 1
customerData[1,1] # Tom with the levels Sallly Sue Tom
# Both info
customerData[1, 1:2] # Tom 43
# All ages
customerData[1:3,2]
# Add record (Record bind)
recordMark = data.frame(name="Mark", age=33)
customerData = rbind(customerData, recordMark)
# Add column (Column bind)
debt = c(0, 25.5, 36, 48.19)
customerData = cbind(customerData, debt)
customerData

# Give customers who owe money
owesMoney = customerData[customerData$debt > 0, ]
owesMoney # Sally Sue and Mark

# head and tail
## head: Shows the beginning of a dataframe
## tail:  //    //   ending  // //   //

head(owesMoney, 2)
tail(owesMoney, 1)
```

# Decision

## Logical operators

```{r}
sprintf("TRUE && FALSE = %s", T&&F) # AND
sprintf("TRUE || FALSE = %s", T||F) # OR
sprintf("!TRUE = %s", !T)           # NOT
```

## Statements

### if statement

```{r}
age = 18

if (age >= 18){
  print("drive and vote")
} else if (age >= 16) {
  print("Drive")
} else {
  print ("wait")
}
```

### Switch

```{r}
grade = "C"

# Waht's after the = get executed
switch(grade,
       "A" = print("Great"),
       "B" = print("Good"),
       "C" = print("Ok"),
       "D" = print("Bad"),
       "F" = print("Terrible"),
       print("No such grade")) # Default
```

# Looping

## Repeat

Until a condition is met
```{r}
num = 1
repeat {
  print(num)
  num = num + 1
  if(num > 5){
    break
  }
}
```

## While

While a condition is true
```{r}
num = 5
while(num > 0) {
  num = num -1
  if (num %% 2 == 0){
    next # Skip the rest of the loop
  }
  print(num) # The odd values
}
```

## For

Cycle through vectors:

```{r}
oneToFive = 1:5
for (i in oneToFive){
  print(i)
}
```

Even shorter:

```{r}
for (i in 1:5){
  print(i)
}
```

# Matrices & Arrays

## Matrices

Store values in rows and columns

```{r}
matrix.1dimension  = matrix(data=c(1,2,3,4))
matrix.2dimensions = matrix(data=c(1,2,3,4), nrow=2, ncol=2) # Automatic

# Fill Rows then columns
matrix.3 = matrix(data = c(1,2,3,4), nrow=2, ncol=2, byrow = T)
dim(matrix.3) # 2 by 2

# Get a value at a ro and a column
matrix.3[1,2]

# Get entire row
matrix.3[1,]

# Get entire column
matrix.3[,2]


# Make a matrix with vectors
## We use rbind

matrix.4 = rbind(1:3, 4:6, 7:9)

# Get the 2 last rows:
matrix.4[2:3,]

# Same result: Omit the first row:
matrix.4[-1,]

# change a value
matrix.4[1,1] = 0

matrix.4[1,] = c(10, 11, 12)
```

## Multi-dimensional arrays

```{r}
# 2 rows, 2 colums, 2 layers
array.3dimensions = array(data = 1:8, dim=c(2,2,2))
array.3dimensions

# Get a value
array.3dimensions[1,2,2]

# ... Everything the same
```

# Functions

In R, a function is just an object that performs operations on passed attributes and then returns the result

## Definition

```{r}
# A basic funciton:
getSum = function(num1, num2){
  return (num1 + num2)
}

sprintf("5 + 6 = %d", getSum(5,6))

# Last statement can be the return:
getDiff = function(num1, num2){
  num1 - num2
}

sprintf("5 - 6 = %d", getDiff(5,6))

# Possible to put default values in signature
getDiv = function(num1, num2=1) { }

# Return multiple value in a list
makelist = function(string){
  return(strsplit(string, " "))
}

# Accept variable number of arg
getSumMore = function (...){
  numList = list(...)
  sum = 0
  for (i in numList){
    sum = sum + i
  }
  return(sum)
}
getSumMore(1,2,4,6,7)
```
## Anonymous function

```{r}
# Anonymous function
numList = 1:10
doubleList =  (function(x) x*2) (numList)
```


Store functions in lists:
```{r}
addFunc = list(
  add2 = function(x) x+2,
  add3 = function(x) x+3
)

addFunc$add2(5)
```

## Clojures

A funciton created by a function

```{r}
power = function(exp){
  function(x){
    x ^ exp
  }
}

cubed = power(3)
cubed(2)   # On a value
cubed(1:5) # On a vector
```

## Exceptions handling

Basic Idea:
```{r}
miss = function(x){
  if (missing(x)) {
    return("Missing argument")
  } else {
    return(x)
  }
}
```

With try catch:
```{r}
divide = function(num1, num2){
  tryCatch(
    num1 / num2,
    error = function(e){
      if(is.character(num1) || is.character(num2)){
        print("Can't divide with strings")
      }
    }
  )
}

divide(2, "5")
divide(2, 5)
```

# I/O

## Read

```{r}
# Pull info in dataframe
myPeople = read.table(file = "./resources/people.txt",
                      header = T,
                      sep = " ",
                      na.strings = "`",
                      stringsAsFactors = F)
myPeople

# Add Donna to the dataframe
## Create Donna
donna = data.frame(fname = "Donna",
                   lname = "Heyward",
                   sex = "female")
## Record-bind her
myPeople = rbind(myPeople, donna)
myPeople
```

## Write

```{r}
# Update record
myPeople[7,2] = "Marx"
write.table(x=myPeople,
            file = "updated_myPeople.txt",
            sep=",", # We create a csv
            na = "`",
            quote = F, # Strings not quoted
            row.names = F)
```

# Plotting

## Scatter plots

```{r}
# First 5: X // last 5 : y
xy1 = matrix(data=c(1, 2, 3, 4, 5, # X
                    1, 2, 3, 4, 5), # Y
             nrow = 5, ncol = 5)
dim(xy1)
plot(xy1)
```

With a line
```{r}
x = c(1,2,3,4,5)
y = c(1,2,3,4,5)
plot(x, y, type="l") # Type "l" as in L-INE
```

With points and line
```{r}
x = c(1,2,3,4,5)
y = c(1,2,3,4,5)
plot(x, y, type="b") # Type "b" as in B-OTH
```

With overriding points and line
```{r}
x = c(1,2,3,4,5)
y = c(1,2,3,4,5)
plot(x, y, type="o") # type "o" as in O-VERRIDE
```

### Labels and title

```{r}
x = c(1,2,3,4,5)
y = c(1,2,3,4,5)
plot(x, y, type="l",
     main="My plot", 
     xlab="x axis", ylab = "y axis",
     col="steelBlue")
```

```{r}
x = c(1,2,3,4,5)
y = c(1,2,3,4,5)
plot(x, y, type="b", pch=2, lty=2, 
     main="My plot", 
     xlab="x axis", ylab = "y axis",
     col="steelBlue",
     xlim=c(-8, 8), ylim=c(-8,8))
```

### Multiple plots

Darw segments:
```{r}
# First plot
x = c(1,2,3,4,5)
y = c(1,2,3,4,5)
plot(x, y, type="b")

# Add a line
abline(h=c(2,4), col="red", lty=2) # Lines @ 2 & 4, red, dotted

# Add segments
segments(x0=c(2,4), y0=c(2,2), x1=c(2,4), y1=c(4,4), col="red")
```

With the built-in data set faithful. Highlight eruptions with a waiting time > than 4:
```{r}
plot(faithful)
eruptions4 = with(faithful, faithful[eruptions > 4,])
points(eruptions4, col = "red", pch = 3)
```

## Draw arrows & text
```{r}

x = c(1,2,3,4,5)
y = c(1,2,3,4,5)
plot(x, y, type="b")
abline(h=c(2,4), col="red", lty=2)
segments(x0=c(2,4), y0=c(2,2), x1=c(2,4), y1=c(4,4), col="red")

# Arrow
arrows(x0=1.5, y0=4.5, x1=2.7, y1=3.25, col="blue")

# Text
text(x=1.25, y=4.75, labels="Center", col="blue")
```

## Pie charts

Food preference for kids

```{r}
foodLabels = c("Spaghetti", "Pizza", "MacNCheese", "Chicken", "Tacos") # Label
foodPref = c(15, 35, 10, 25, 15) # Preference
#png(file="childfoodPreference.png")
colors = rainbow(length(foodPref))
pie(foodPref, foodLabels, main="Food Prefs", col=colors)
legend("topright", c("Spaghetti", "Pizza", "MacNCheese", "Chicken", "Tacos"), cex=0.8,
       fill=colors)
```

## 3D Pie chart

We have to install a package. In te console:

```sh
> install.packages("plotrix")
```

```{r}
# Load library
library(plotrix)

# Prepre
foodLabels = c("Spaghetti", "Pizza", "MacNCheese", "Chicken", "Tacos") # Label
foodPref = c(15, 35, 10, 25, 15) # Preference

# Plot
pie3D(foodPref, labels="foodLabels", main="Food Prefs", explode=0.05, start=pi/2, labelcex=0.8)
```

## Bar charts

```{r}
# Prepre
foodLabels = c("Spaghetti", "Pizza", "MacNCheese", "Chicken", "Tacos") # Label
foodPref = c(15, 35, 10, 25, 15) # Preference
colors = rainbow(length(foodPref))

# Plot
barplot(foodPref, name="foodLabels", main="Food Prefs", xlab="Votes", ylab="food option", col=colors)
```

# Math functions

## Basics

```{r}
sqrt(x=4)
log(x=4, base=2)
exp(x=2)
sum(c(1,2,3))
random = c(1,5,6,7,10,16)
mean(random)
median(random)
min(random)
max(random)
range(random) # min and max at the same time
ceiling(4.5)
round(4.5)
```

## More advanced

```{r}
cumsum(c(1,2,4)) # Add 1 to 1, 2 to 1, 4 to 2
cumprod(c(1,3,7)) # same with *
cummax(c(7:9, 4:8, 1:10)) # Cummulate and sticks to the max it finds as it loops
cummin(c(7:9, 4:8, 1:3))
```

## Random

```{r}
# Coin flip
sample(0:1, 10, replace=T) # Replace means weigh the probability of the next flip based on the previou
```

# Regression analysis

Study correlation between two pieces of data.
```{r}
mlbPlayers = read.table(file="./resources/mlb2017players.txt", 
                        header=T, 
                        sep=" ",
                        na.strings="`",
                        stringsAsFactors=F)
playerData = mlbPlayers[,c("RBI", "AVG")]
relation =lm(playerData$RBI~playerData$AVG)
plot(playerData$AVG, playerData$RBI, main="AVG & RBI regression", abline(lm(playerData$RBI~playerData$AVG)), xlab="AVG", ylab="RBIs")
```

